Assessing reliability of an online behavioural laterality battery: a pre-registered

### 2020-07-10 draft_notebook_writeup.Rmd
full first draft

## Authors
Adam J. Parker, Zoe V. J., Woodhead, Paul A. Thompson, and Dorothy V. M. Bishop

@@

<!--- Leaving blank for now as not sure how corresponding authorship works in OSCCI lab --->
*Correspondence regarding this article should be addressed to XXXXXX, Department of Experimental Psychology, University of Oxford, Anna Watts Building, Woodstock Road, Oxford, OX2 6GG. Email: XXXXXX

## Abstract

**Needs filling in**

## Introduction

Hemispheric specialisation, or lateralisation, is a ubiquitous characteristic of the human brain, yet many questions remain unanswered as to its causes and consequences. Much of the ambiguity in the published literature may arise from unreliable measures of lateralisation and underpowered studies. Without well powered experiments that use reliable tasks, it becomes difficult to address many research questions regarding the relationship between laterality indices and their stability over time. With respect to the relationship between indices, unless a task is reliable, the absence of a correlation cannot be interpreted as evidence for independence of function (Bishop, 2017). The best way to determine a task’s reliability is to test it twice. Here, the variation between participants should be larger than the variation within subjects. Taking this into consideration, the current work introduces an online behavioural laterality battery for which we established the test-retest reliability using a sample of 390 participants. The behavioural tasks included were: a novel rhyme decision task, a dichotic listening task (Hugdahl et al., 2009), a chimeric faces task (Burt & Perret, 1997), and a novel finger tapping task. We subsequently use this battery to we examine two pre-registered questions: (1) do extreme left-handers show a higher rate of atypical laterality than right handers on each behavioural laterality task; (2) do the laterality indices produced on each measure correlate.

@@

Behavioural laterality tasks have long been used to establish hemispheric dominance for certain cognitive functions. Rather consistently, visual half-field and dichotic listening paradigms agree that the left hemisphere plays the dominant role in language processing  (Hugdhal, 2000; Ocklenburg & Güntürkün, 2018). Here we review how each paradigm has been implemented to study language lateralisation.

Evidence from the visual half-field paradigm (see Bourne, 2006, for a review), strongly points towards advantageous processing for letter strings presented in the right, relative to the left, visual field. As information in the right visual field projects contralaterally and arrives directly at the language dominant left hemisphere, it is thought to reflect an absence in callosal transfer that is required for the processing of information initially arriving at the right hemisphere.

Despite wide reporting of a right visual field advantage at the population level when using visual half-field tasks, the paradigm is not without critique. In particular, the validity of visual half-field tasks has been questioned. Krach, Chen, and Hartje (2006) compared the laterality indices (LIs) produced during a visual half-field experiment with those produced using functional Transcranial Doppler Sonography (fTCD). The visual half-field experiment itself consisted of lexical decisions made towards abstract nouns presented in the left and right visual-fields. During the experiment, the blood flow to left and right hemispheres of 58 participants was measured. In addition, blood flow was measured during a word generation task—the gold standard task for assessing hemispheric dominance for language. Despite showing the expected right visual field advantage on the visual half-field task, the behavioural LIs produced on the visual half-field paradigm did not correlate with the fTCD LIs produced during word generation (r= .18) or the visual half-field task itself (r -.004). The 
lack of validity here likely reflects the poor reliability of the visual field task as Krach et al. report the split-half reliability for the correct response rates and reaction times during the behavioural visual half-field paradigm were r= .14 and r= .20 respectively. Thus, with such low split-half reliability it is unclear whether the task was able to accurately measure language lateralisation to start with.

In recent years it has become clear that the reliability and validity of visual half-field tasks to identify the language-dominant hemisphere is largely dependent on several methodological aspects. For example, Hunter and Brysbaert (2008) noted stronger and more stable visual-field differences when stimuli were presented bilaterally relative to those where only one stimulus was presented either in the LVF or in the RVF (Boles, 1987, 1990, 1994; see also Iacoboni & Zaidel, 1996). Brysbaert and Hunter argued that this is the result of stimuli competing during bilateral presentation, which is easier when the stimulus is presented in the language dominant hemisphere. Reliability and validity also depend on there being a sufficient number of observations and brief stimulus presentation. When accounting for these methodological aspects, the validity of visual half-field tasks seems promising. For example, Hunter and Brysbaert (2008) reported that LIs from visual half-field paradigms correlated positively with those from functional Magnetic Resonance Imaging (fMRI): word naming (r= .63), picture naming (r= .77). Here 192 trials were used during word naming and 160 trials during picture naming while stimuli were bilaterally presented and masked at offset. Furthermore, a promising visual half-field paradigm developed by Hausmann and colleagues (Hausmann et al., 2019; Willemin et al., 2016) which adhered to several of the methodological aspects outline by Brysbaert and Hunter has reported a right visual-field advantage at the population level with little variation occurring between languages. The authors, however, have not yet reported reliability measures for this task.

While the visual half-field technique measures lateralisation for visual stimuli, dichotic listening tasks examine hemispheric dominance for the processing of auditory stimuli. Like the visual half-field technique, the dichotic listening task makes use of the crossing of fibres at the corpus callosum. In this task, different streams of auditory linguistic stimuli (i.e. words, syllables) are presented to each ear and participants are either asked simply to report as many words as they can or to attend to one ear selectively. In either case, a right ear advantage has been demonstrated; typically, participants report more verbal stimuli presented to the right ear (Hugdahl, 2011; Kimura, 1961). Again, this advantage has been attributed to stimuli being projected to the language dominant hemisphere without callosal transfer.

Over the years, various dichotic listening paradigms have been used (Bryden, 1988; Westerhausen, 2019). The test-retest reliability of these paradigms have been far from optimal (Voyer, 1998; Kelley & Littenberg, 2019). However, recent data from Westerhausen and Samuelsen (2020) indicated that when designed well, the dichotic listening paradigm can be reliable. First, based on a review of the published literature, Westerhausen and Samuelsen outlined eight design features that should be used in an optimal dichotic listening paradigm and explain their justification for each. The criteria are as follows: (1) stop consonant-vowel (CV) syllables as stimulus materials, (2) pair only CV stimuli from the same voicing category, (3) Alternating trials of voiced and unvoiced stimulus pairs, (4) free-recall instruction, (5) single-stimulus pair per trial, (6) paradigm length of 120 dichotic trials, (7) all stimulus pairs are presented in both orientations in identical frequency, and (8) includes binaural (homonymic) trials. Putting these criteria into practice, Westerhausen and Samuelsen administered a modified dichotic listening paradigm using both a verbal and manual response format to 50 right-handers. Interestingly, the full paradigm yielded a test-retest reliability of r= .93 for verbal responses and r= .91 for manual responses. Thus, the dichotic listening task provides a promising paradigm for those wanting to investigate language lateralisation without reliance on visual stimuli.

@@

Of importance to the current work is the use of the visual half-field paradigm to measure the lateralisation of face emotionality, in particular the chimeric faces task. In this task participants are presented with vertically split chimeric faces, where each half displays a different emotion. Depending on how the task is implemented, participants will either view a single chimeric face and report the emotion being expressed (e.g. Drebing, Federman, Edington, & Terzian, 1997; Kucharska-Pieture, & David, 2003) or view two faces, one above the other, and decide which of the two faces looks more emotive (e.g., Bourne, 2005; Levy, Heller, Banich, & Burton, 1983). Here the images will be identical, mirror images and typically include a neutral and emotive hemiface (i.e. Happiness-Neutral vs Neutral-Happiness). Regardless of implementation, participants’ responses point towards a bias for the processing of emotion in the left visual field. In the single face version, participants will more often report the emotion shown in the left hemiface. In the two-face version, participants show a bias in selecting the face with an emotion depicted in the left hemiface.

In terms of its reliability, the chimeric faces tasks seem promising. In Levy et al.’s (1983) early study, they reported good split-half reliability for the two faces version of the chimeric faces task, where the left and right handers both showed a split-half reliability of .93. The validity of the chimeric faces task has been demonstrated using behavioural and fMRI data.  In Yovel, Tambini, and Brandman’s (2008) study, participants performed a behavioural experiment where they had to recognize centrally presented chimeric faces, which presented different identities in the right- and left-visual-field. Interestingly, the behavioural measure significantly correlated with the LI calculated based on activation in the fusiform face area (r= .49). However, it is of importance to note the sample size used here was small, but at face value the results seem promising.

@@

The studies reviewed so far have all focused on lateralisation at the population level. That is not to say that lateralisation does not vary between individuals. One factor that has been associated with the degree of lateralisation is handedness. Neuroimaging data strongly indicates that these differences in lateralisation are the result of a group of left-handers who are right language dominant rather than inidciateing that this relationship operates on a continuum (Mazoyer et al., 2014This increased departure from typical lateralisation in left-handers is not only present in neuroimaging studies, it is also reflected in behavioural tasks. Thus it will be important for our battery to replicate the findings regarding handedness and lateralisation summarised below.

With regards to visual half-field techniques for studying language lateralisation, there is evidence to suggest that the visual field advantage varies with handedness (Elias, Bulman-Fleming, &McManus, 1999; Nicholls, 1994, 1998; Weekes, Capetillo-Cuncliffe, Rayman, Iacoboni, & Zaidel, 1999). However, a relatively well powered study, involving 100 participants conducted by Willemin et al. (2016) reported no effect of handedness (when coded as left and right) on the visual field advantaged during lexical decision task. At first, this finding may seem odd. However, it may reflect underlying issues with the task rather than represent strict evidence for a null effect of handedness on visual half-field task. Here, Willemin et al. asked participants to respond via manual response. This contrasts several studies that have reported increased variability when responding aloud. Thus, a task that lacks some active speech production component may not yield such strong visual field advantages that enable provide enough variation to detect subtle differences in handedness (Hunter & Brysbaert, 2008). This, of course, is something that we kept in mind when developing the visual half-field paradigm used in the current study.

When dichotic listening paradigms to study language lateralisation, the evidence is clear that non-right-handers differ in terms of their laterality. While a clear right ear (i.e. left hemisphere) advantage is noted at the population level (see Hugdahl et al., 2009, and Westerhausen, 2019, for a review), there is evidence to suggest that this is reduced in left- relative to right-handers (Bethmann, Tempelmann, De Bleser, Scheich, & Brechmann, 2007; Bryden, 1970, 1988; Hugdahl, 2003); although handedness alone does not predict this difference (Van der Haegen, Westerhausen, Hugdahl, & Brysbaert, 2013). Across two studies Karlsson, Johnstone, and Carey (2019) provided a clear, yet comprehensive explanation of how performance on the dichotic listening task differed as a function of hand preference. Here, Karlsson et al. employed a consonant-vowel version of the dichotic listening task where participants reported which one of six syllables that they heard the clearest. For study 1, a clear right ear advantage was reported at the population level. Furthermore, right-handers were on average more lateralised than non-right-handers—87% of right-handers and 78% of non-right-handers showed a right ear advantage. For study 2, a similar pattern of results emerged. Again, right-handers were more lateralised with 85% of right-handers and 79% of non-right-handers showing an advantage. Together these results indicate that non-right-handers show a larger departure from the ‘typical’ lateralisation during dichotic listening tasks.

@@

From the above reviewed studies, it appears that handedness is related to laterality across a range of tasks. This relationship points towards a a general laterality factor which governs lateralisation across tasks and behaviours. This has been argued in several hypotheses (Behrmann & Plaut, 2015; Cook, 1984; Hellige, 1983; Kosslyn, 1987). Common to all hypotheses is the assumption that there would be a correlation between the degree of laterality across complementary tasks. While studies do report significant correlations between measures of lateralisation, these are often small (Bradzakova-Trajkov, Haberling, Roberts, & Corballis, 2010). Furthermore, there are a large number of studies which report no correlation between laterality indices (Bryden, Hecaen, & Deagostini, 1983; Rosch, Bishop, & Badcock, 2012; Whitehouse & Bishop, 2009; van der Haegen & Brysbaert, 2018). This has led to the conclusion that there are multiple independent biases across tasks which lead to hemispheric lateralisation of different cognitive domains.

@@

Our second prediction related to the inter-relationships between laterality indices. With online testing, there is the opportunity to test inter-relationships between laterality indices with an adequately powered sample using measures of known reliability.  Prior studies have been inconclusive because it has been difficult gather large amounts of data to establish test-retest reliability. For instance, Van der Haegen and Brysbaert (2018), found that language laterality did not correlate with face laterality in visual half-field tasks. However, their study was restricted to left-handers, who may be atypical in this regard, and they noted that lack of association was hard to interpret because it may be due to unreliability of measurement. Originally, we predicted that for right-handers, a model (a) where the two language tasks are correlated with each other but not with the chimeric faces will provide better fit to the data than a model (b) in which laterality indices are correlated between the language and chimeric face tasks, or a base model (c) where each task is independently lateralised. For left-handers, we anticipated that that model (c) may be a better fit to the data, indicating that laterality is dissociated across the different tasks. However, we were not able to test this planned analysis with the current data set as one task had low reliability. Thus, we instead examined the relationship between laterality tasks using correlation analysis.

Our final prediction that we examined relates to handedness and atypical lateralisation. We predicted that strong left-handers (i.e. those with laterality quotient of -90 or less on the Edinburgh Handedness Inventory) will be more likely to show atypical laterality (i.e. laterality in the opposite direction to our population level predictions) than right-handers. This prediction follows from observations by Mazoyer et al (2014, p. 1) who studied a large group of left- and right-handers using fMRI to measure language laterality. They concluded "concordance of hemispheric dominance for hand and for language occurs barely above the chance level, except in a group of rare individuals (less than 1% in the general population) who exhibit strong right hemisphere dominance for both language and their preferred hand”.

@@

As our ultimate goal is to use these measures in individual differences research, we aimed to optimise the task to achieve observed test-retest reliability of at least .65. To achieve sufficient precision of the estimate of reliability, we estimated we needed 300 participants. With this sample size, if the true correlation is .58, then the 95% confidence interval around the estimate will be .50-.65. Thus with this sample size, if we observe a correlation of .65, then the true correlation is unlikely to be lower than .58.

Out of the 300 participants, we originally specified that we'd collected data for least 80 left-handers. We had examined the power that a sample of 80 participants would provide to make the originally planned model comparisons outlined under hypothesis 3. By comparing model fit using a likelihood test, 80 participants would provide 100% power for model (a) and 97% power for model (b). Thus, we would have sufficient power to assess the fit of each model using Akaike weights, which are more conservative than likelihood tests.

@@

<!--Dorothy, just a note that all Ps came from Prolific in the end-->

@@

The basic demographics questionnaire asked participants to report handedness, years in education, and whether they were bilingual. We also asked participnats a question about footedness: which foot do you normally use to step up on a ladder/step? Participants could answer "left", "right", or "don't know".

@@

The Edinburgh Handedness Inventory (Oldfield, 1971) was administered to quantify handedness on a continuum. For 10 everyday activities, participants indicated their preferred hand use on a 5-step scale (right hand strongly preferred, right hand preferred, no preference, left hand preferred, left hand strongly preferred). The Edinburgh handedness inventory was scored in the standard way, which involves combining information about direction and exclusiveness of hand preference across ten tasks. For simple contrast between left- and right-handers, we treated indices greater than 0 as right handed, and indices less than 0 as left handed. For analysis of extreme left-handers, we focussed on left-handers with an index of -90 or less, which corresponds to near-exclusive use of the left hand.

@@

The rhyme decision task is a novel visual half-field technique which requires participants decide whether a foveally present letter string rhymed with an image shown in either the left or right hemifield.

@@

Each trial began with a fixation cross presented centrally for 800 ms. Participants were instructed to keep their eyes on the central fixation point. Then a foveally presented word would appear for 200 ms. Word stimuli were presented in black lowercase Tahoma font on a white background. Font was scaled to 250% equating to 28pt font. The foveally presented word was replaced with bilaterally presented images at approximately 2.5 of visual angle either side of fixation. The eccentricity of images was maintained using Gorilla’s scaling tool. The bilaterally presented images were then presented for 150 ms. Brief bilateral presentation has guaranteed adequate control for eye movements in previous VHF experimental with verbal naming (Beaumont, 1982). Studies that directly monitored eye movements reported failures of fixation on only 0.5% of trials (Geffen, Bradshaw & Nettleton, 1972, but see also Bourne, 2006). At stimulus offset, participants indicated whether the image in the LVF or RVF rhymed with the foveally presented word. Participants were instructed to press ‘S’ to indicate that the foveally presented word rhyme with the image in the LVF and ‘K’ if it rhymed with the image in the RVF. If participants did not detect a rhyme, they were instructed to press the space bar. The response triggered the next trial.

@@

#### General Procedure

@@

#### Data analysis

@@

##### Calculations of laterality indices

For the rhyme task, we pre-registered that we would calculate LIs based on reaction time data, as accuracy during piloting was at ceiling. We specified that we would calculate LIs in two ways: (1) LIs would be calculated in the traditional fashion (100 x (Right – Left)/(Right + Left)); (2) LIs would be calculated in a way that corresponded to a *t*-value. This was achieved by conducting a *t*-test for each participant where accuracy response times are the dependent variable and visual field is the independent variable. This would estimate the sensitivity to stimuli presented in either visual field and equate to a laterality index. One benefit of this approach is that we could identify participants who show a statistically significant reaction time advantage for one side, i.e. where the *t*-value on a 2-sided test has *p*< .05. Based on considerations from pilot data, we planned to transform RT data from the RDT task by log-transformation before computing *t*-values.

@@

##### Reliability assessment

For the reliability assessment, we plotted the distributions of LIs and checked them for normality using a Shapiro-Wilk test. As the LIs were non-normal, we used a series of Spearman’s correlations to examine the correlation of coefficients for session 1 versus session 2.

For tasks where the test-retest reliability was above .75 with the existing test length, we reanalysed the data using just the first N trials, reducing N until reliability becomes unsatisfactory (below *r*= .75)^1^. This will allow us to identify the most efficient way to administer these tasks in future, while retaining good reliability. However, the full data (including all trials) was used for all subsequent analyses.

##### Population level analysis

To address hypothesis 1 and examine the laterality of each task at the population level, we conducted a series of one-sample *t*-tests for each task using session 1 data.

>>>>

Adam J. Parker

Zoe V. J., Woodhead

Paul A. Thompson

Dorothy V. M. Bishop

@@

<!--- not sure how corresponding authorship works in OSCCI lab --->
*Correspondence regarding this article should be addressed to Adam J. Parker, Department of Experimental Psychology, University of Oxford, Anna Watts Building, Woodstock Road, Oxford, OX2 6GG. Email: adam.parker@psy.ox.ac.uk.

\newpage
## Abstract

Studies of cerebral lateralisation often involve participants completing a series of perceptual tasks under laboratory conditions. This has constrained the number of participants recruited in such studies. The current work, however, introduces an online laterality battery that can be used to test participants on a variety of tasks: a rhyme decision visual half-field task, a dichotic listening task, a chimeric faces task, and a finger tapping task to assess manual preference. To validate the battery and assess its test-retest reliability, 392 participants completed the battery twice. Around a third of participants were left-handers. Not only do we report hemispheric advantages in the expected direction for each task, we additionally report good test-retest for the dichotic listening, chimeric faces, and finger tapping tasks. We also examine the association between handedness and atypical lateralisation and report that left-handers are more likely to depart from typical lateralisation. Furthermore, we provide data which indicates that lateralisation across tasks likely stems from multiple independent sources. The implications of these findings are discussed.

*Keywords*: cerebral lateralisation, hemispheric asymmetry, language, online behaviour research methods, reliability

\newpage
## Introduction

Hemispheric specialisation, or lateralisation, is a ubiquitous characteristic of the human brain, yet many questions remain unanswered as to its causes and consequences. Much of the ambiguity in the published literature may arise from unreliable measures of lateralisation and underpowered studies. Without well powered experiments that use reliable tasks, it becomes difficult to address many research questions regarding the relationship between laterality indices and their stability over time. With respect to the relationship between indices, unless a task is reliable, the absence of a correlation cannot be interpreted as evidence for independence of function (Bishop, 2017). The best way to determine a task’s reliability is to test it twice. Here, the variation between participants should be larger than the variation within subjects. Taking this into consideration, the current work introduces an online behavioural laterality battery for which we established the test-retest reliability using a sample of 392 participants. The behavioural tasks included were: a novel rhyme decision task, a dichotic listening task (Hugdahl et al., 2009), a chimeric faces task (Burt & Perret, 1997), and a novel finger tapping task. We subsequently use this battery to we examine two pre-registered questions: (1) do extreme left-handers show a higher rate of atypical laterality than right handers on each behavioural laterality task; (2) do the laterality indices produced on each measure correlate.

@@

Behavioural laterality tasks have long been used to establish hemispheric dominance for certain cognitive functions. Rather consistently, results from visual half-field and dichotic listening paradigms suggest that the left hemisphere plays the dominant role in language processing (Hugdhal, 2000; Ocklenburg & Güntürkün, 2018). Here we review how each paradigm has been implemented to study language lateralisation and each paradigms’ reliability.

Evidence from the visual half-field paradigm (see Bourne, 2006, for a review), strongly points towards advantageous processing for letter strings presented in the right, relative to the left, visual field. As information in the right visual field projects contralaterally and arrives directly at the language dominant left hemisphere, it is thought to reflect an absence in callosal transfer that is required for the processing of linguistic information initially arriving at the right hemisphere.

Despite wide reporting of a right visual field advantage at the population level when using visual half-field tasks, the paradigm is not without critique. In particular, the validity of visual half-field tasks has been questioned. Krach, Chen, and Hartje (2006) compared the laterality indices (LIs) produced during a visual half-field experiment with those produced using functional Transcranial Doppler Sonography (fTCD). The visual half-field experiment itself consisted of lexical decisions made towards abstract nouns presented in the left- and right visual-fields. During the experiment, the blood flow to left and right hemispheres of 58 participants was measured. In addition, blood flow was measured during a word generation task—the gold standard task for assessing hemispheric dominance for language. Despite showing the expected right visual field advantage on the visual half-field task, the behavioural LIs produced on the visual half-field paradigm did not correlate with the fTCD LIs produced during word generation (r= .18) or the visual half-field task itself (r -.004). The lack of validity here likely reflects the poor reliability of the visual field task as split-half reliability was far from optimal (RT spilt-half: r= .14; accuracy split-half: r= .20). Thus, with such low split-half reliability it is unclear whether the task was able to accurately measure language lateralisation to start with.

In recent years it has become clear that the reliability and validity of visual half-field tasks to identify the language-dominant hemisphere is largely dependent on several methodological aspects. For example, Hunter and Brysbaert (2008) noted stronger and more stable visual-field differences when stimuli were presented bilaterally relative to those where only one stimulus was presented either in the LVF or in the RVF (Boles, 1987, 1990, 1994; see also Iacoboni & Zaidel, 1996). Brysbaert and Hunter argued that this is the result of stimuli competing during bilateral presentation, which is easier when the stimulus is presented in the language dominant hemisphere. Reliability and validity also depend on there being a sufficient number of trials and brief stimulus presentation. When accounting for these methodological aspects, the validity of visual half-field tasks seems promising. For example, Hunter and Brysbaert (2008) reported that LIs from visual half-field paradigms correlated positively with those from functional Magnetic Resonance Imaging (fMRI): word naming (r= .63), picture naming (r= .77). Here 192 trials were used during word naming and 160 trials during picture naming while stimuli were bilaterally presented and masked at offset. Furthermore, a promising visual half-field paradigm developed by Hausmann and colleagues (Hausmann et al., 2019; Willemin et al., 2016) which adhered to several of the methodological aspects outline by Brysbaert and Hunter has reported a right visual-field advantage at the population level with little variation occurring between languages. The authors, however, have not yet reported reliability measures for this task.

While the visual half-field technique measures lateralisation for visual stimuli, dichotic listening tasks examine hemispheric dominance for the processing of auditory stimuli. Like the visual half-field technique, the dichotic listening task makes use of the crossing of fibres at the corpus callosum. In this task, different streams of auditory linguistic stimuli (i.e. words, syllables) are presented to each ear and participants are either asked simply to report as many words as they can or to attend to one ear selectively. In either case, a right ear advantage has been demonstrated; participants report more verbal stimuli presented to the right ear (Hugdahl, 2011; Kimura, 1961). Again, this advantage has been attributed to stimuli being projected to the language dominant hemisphere without callosal transfer.

Over the years, various dichotic listening paradigms have been used (Bryden, 1988; Westerhausen, 2019). As with the visual half-field paradigm, the test-retest reliability of dichotic listening paradigms has been far from optimal (Voyer, 1998; Kelley & Littenberg, 2019). However, recent data from Westerhausen and Samuelsen (2020) indicated that when designed well, the dichotic listening paradigm can be reliable. First, based on a review of the published literature, Westerhausen and Samuelsen outlined eight design features that should be used in an optimal dichotic listening paradigm and explain their justification for each. The criteria are as follows: (1) stop consonant-vowel (CV) syllables as stimulus materials, (2) pair only CV stimuli from the same voicing category, (3) Alternating trials of voiced and unvoiced stimulus pairs, (4) free-recall instruction, (5) single-stimulus pair per trial, (6) paradigm length of 120 dichotic trials, (7) all stimulus pairs are presented in both orientations in identical frequency, and (8) includes binaural (homonymic) trials. Putting these criteria into practice, Westerhausen and Samuelsen administered a modified dichotic listening paradigm using both a verbal and manual response format to 50 right-handers. Interestingly, the full paradigm yielded a test-retest reliability of r= .93 for verbal responses and r= .91 for manual responses. Thus, the dichotic listening task provides a promising paradigm for those wanting to investigate language lateralisation without reliance on visual stimuli.

@@

Of importance to the current work is the use of the visual half-field paradigm to measure the lateralisation of face emotionality, in particular the chimeric faces task. In this task participants are presented with vertically split chimeric faces, where each half displays a different emotion. Depending on how the task is implemented, participants will either view a single chimeric face and report the emotion being expressed (e.g. Drebing, Federman, Edington, & Terzian, 1997; Kucharska-Pieture, & David, 2003) or view two faces, one above the other, and decide which of the two faces looks more emotive (e.g., Bourne, 2005; Levy, Heller, Banich, & Burton, 1983). In the two face version images will be identical, mirror images and typically include a neutral and emotive hemiface (i.e. Happiness-Neutral vs Neutral-Happiness). Regardless of implementation, participants’ responses point towards a bias for the processing of emotion in the left visual field. In the single face version, participants will more often report the emotion shown in the left hemiface. In the two-face version, participants show a bias in selecting the face with an emotion depicted in the left hemiface.

In terms of its reliability, the chimeric faces tasks seem promising. In Levy et al.’s (1983) early work, they reported good split-half reliability for the two faces version of the chimeric faces task, where the left and right handers both showed a split-half reliability of .93. The validity of the chimeric faces task has been demonstrated using behavioural and fMRI data. In Yovel, Tambini, and Brandman’s (2008) study, participants performed a behavioural experiment where they had to recognize centrally presented chimeric faces, which presented different identities in the right- and left-visual-field. Interestingly, the behavioural measure significantly correlated with the LI calculated based on activation in the fusiform face area (r= .49). However, it is of importance to note the sample size used here was small, but at face value the results seem promising.

@@

The studies reviewed so far have all focused on lateralisation at the population level. That is not to say that lateralisation does not vary between individuals. One factor that has been associated with the degree of lateralisation is handedness. Neuroimaging data strongly indicates that these differences in lateralisation are the result of a group of left-handers who are right language dominant rather than indicating that this relationship operates on a continuum (Mazoyer et al., 2014). This increased departure from typical lateralisation in left-handers is not only present in neuroimaging studies, it is also reflected in behavioural tasks. Thus it will be important for our battery to replicate the findings regarding handedness and lateralisation summarised below.

With regards to visual half-field techniques for studying language lateralisation, there is evidence to suggest that the visual field advantage varies with handedness (Elias, Bulman-Fleming, & McManus, 1999; Nicholls, 1994, 1998; Weekes, Capetillo-Cuncliffe, Rayman, Iacoboni, & Zaidel, 1999). However, a relatively well powered study, involving 100 participants conducted by Willemin et al. (2016) reported no effect of handedness (when coded as left and right) on the visual field advantaged during lexical decision task. At first, this finding may seem odd. However, it may reflect underlying issues with the task rather than represent strict evidence for a null effect of handedness on visual half-field task. Here, Willemin et al. asked participants to respond via manual response. This contrasts several studies that have reported increased variability when responding aloud. Thus, a task that lacks some active speech production component may not yield such strong visual field advantages that enable provide enough variation to detect subtle differences in handedness (Hunter & Brysbaert, 2008). This, of course, is something that we kept in mind when developing the visual half-field paradigm used in the current study.

When using dichotic listening paradigms to study language lateralisation, the evidence is clear that non-right-handers differ in terms of their laterality. While a clear right ear (i.e. left hemisphere) advantage is noted at the population level (see Hugdahl et al., 2009, and Westerhausen, 2019, for a review), there is evidence to suggest that this is reduced in left- relative to right-handers (Bethmann, Tempelmann, De Bleser, Scheich, & Brechmann, 2007; Bryden, 1970, 1988; Hugdahl, 2003); although handedness alone does not predict this difference (Van der Haegen, Westerhausen, Hugdahl, & Brysbaert, 2013). Across two studies Karlsson, Johnstone, and Carey (2019) provided a clear, yet comprehensive explanation of how performance on the dichotic listening task differed as a function of hand preference. Here, Karlsson et al. employed a consonant-vowel version of the dichotic listening task where participants reported which one of six syllables that they heard the clearest. For study 1, a clear right ear advantage was reported at the population level. Furthermore, right-handers were on average more lateralised than non-right-handers—87% of right-handers and 78% of non-right-handers showed a right ear advantage. For study 2, a similar pattern of results emerged. Again, right-handers were more lateralised with 85% of right-handers and 79% of non-right-handers showing an advantage. Together these results indicate that non-right-handers show a larger departure from the ‘typical’ lateralisation during dichotic listening tasks.

@@

From the above reviewed studies, it appears that handedness is related to laterality across a range of tasks. This relationship points towards a general laterality factor which governs lateralisation across tasks and behaviours. This has been argued in several hypotheses (Behrmann & Plaut, 2015; Cook, 1984; Hellige, 1983; Kosslyn, 1987). Common to all hypotheses is the assumption that there would be a correlation between the degree of laterality across complementary tasks. While studies do report significant correlations between measures of lateralisation, these are often small (Bradzakova-Trajkov, Haberling, Roberts, & Corballis, 2010). Furthermore, there are a large number of studies which report no correlation between laterality indices (Bryden, Hecaen, & Deagostini, 1983; Rosch, Bishop, & Badcock, 2012; Whitehouse & Bishop, 2009; van der Haegen & Brysbaert, 2018). This has led to the conclusion that there are multiple independent biases across tasks which lead to hemispheric lateralisation of different cognitive domains.

@@

Our second prediction that we examined relates to handedness and atypical lateralisation. We predicted that strong left-handers (i.e. those with laterality quotient of -90 or less on the Edinburgh Handedness Inventory) will be more likely to show atypical laterality (i.e. laterality in the opposite direction to our population level predictions) than right-handers. This prediction follows from observations by Mazoyer et al (2014, p. 1) who studied a large group of left- and right-handers using fMRI to measure language laterality. They concluded “concordance of hemispheric dominance for hand and for language occurs barely above the chance level, except in a group of rare individuals (less than 1% in the general population) who exhibit strong right hemisphere dominance for both language and their preferred hand”.

Our final prediction related to the inter-relationships between laterality indices. With online testing, there is the opportunity to test inter-relationships between laterality indices with an adequately powered sample using measures of known reliability. Prior studies have been inconclusive because it has been difficult gather large amounts of data to establish test-retest reliability. For instance, Van der Haegen and Brysbaert (2018), found that language laterality did not correlate with face laterality in visual half-field tasks. However, their study was restricted to left-handers, who may be atypical in this regard, and they noted that lack of association was hard to interpret because it may be due to unreliability of measurement. Originally, we predicted that for right-handers, a model (a) where the two language tasks are correlated with each other but not with the chimeric faces will provide better fit to the data than a model (b) in which laterality indices are correlated between the language and chimeric face tasks, or a base model (c) where each task is independently lateralised. For left-handers, we anticipated that that model (c) may be a better fit to the data, indicating that laterality is dissociated across the different tasks. However, we were not able to test this planned analysis with the current data set as one task had low reliability. Thus, we instead examined the relationship between laterality tasks using correlation analysis.

@@

As our ultimate goal is to use these measures in individual differences research, we aimed to optimise the task to achieve observed test-retest reliability of at least .65. To achieve sufficient precision around the estimate of reliability, we estimated that we would require 300 participants. Out of the 300 participants, we originally specified that 80 would be left-handers. We had examined the power that a sample of 80 participants would provide to make the originally planned model comparisons outlined under hypothesis 2. By comparing model fit using a likelihood test, 80 participants would provide 100% power for model (a) and 97% power for model (b). Thus, we would have sufficient power to assess the fit of each model using Akaike weights, which are more conservative than likelihood tests.

@@

@@

The basic demographics questionnaire asked participants to report handedness, years in education, and whether they were bilingual. We also asked participants a question about footedness: which foot do you normally use to step up on a ladder/step? Participants could answer "left", "right", or "don't know".

@@

The Edinburgh Handedness Inventory (Oldfield, 1971) was administered to quantify handedness on a continuum. For 10 everyday activities, participants indicated their preferred hand use on a 5-step scale (right hand strongly preferred, right hand preferred, no preference, left hand preferred, left hand strongly preferred). The Edinburgh handedness inventory was scored in the standard way, which involves combining information about direction and exclusiveness of hand preference. For simple contrast between left- and right-handers, we treated indices greater than 0 as right handed, and indices less than 0 as left handed. For analysis of extreme left-handers, we focussed on left-handers with an index of -90 or less, which corresponds to near-exclusive use of the left hand.

@@

The rhyme decision task is a novel visual half-field technique which requires participants to decide whether a foveally presented letter string rhymed with an image shown in either the left or right hemifield.

@@

Each trial began with a fixation cross presented centrally for 800 ms. Participants were instructed to keep their eyes on the central fixation point. Then a foveally presented word would appear for 200 ms. Word stimuli were presented in black lowercase Tahoma font on a white background. Font was scaled to 250% equating to 28pt font. The foveally presented word was replaced with bilaterally presented images at approximately 2.5 deg. of visual angle either side of fixation. The eccentricity of images was maintained using Gorilla’s scaling tool. The bilaterally presented images were then presented for 150 ms. Brief bilateral presentation has guaranteed adequate control for eye movements in previous VHF experimental with verbal naming (Beaumont, 1982). Studies that directly monitored eye movements reported failures of fixation on only 0.5% of trials (Geffen, Bradshaw & Nettleton, 1972, but see also Bourne, 2006). At stimulus offset, participants indicated whether the image in the LVF or RVF rhymed with the foveally presented word. Participants were instructed to press ‘S’ to indicate that the foveally presented word rhyme with the image in the LVF and ‘K’ if it rhymed with the image in the RVF. If participants did not detect a rhyme, they were instructed to press the space bar. The response triggered the next trial.

@@

### General Procedure

@@

### Data analysis

@@

##### Calculations of laterality indices, reliability, and population level analysis

For the rhyme task, we pre-registered that we would calculate LIs based on reaction time data, as accuracy during piloting was at ceiling. We specified that we would calculate LIs in two ways: (1) LIs would be calculated in the traditional fashion (100 x (Right – Left)/(Right + Left)); (2) LIs would be calculated in a way that corresponded to a *t*-value. This was achieved by conducting a *t*-test for each participant where accurate response times are the dependent variable and visual field is the independent variable. This would estimate the sensitivity to stimuli presented in either visual field and equate to a laterality index. One benefit of this approach is that we could identify participants who show a statistically significant reaction time advantage for one side, i.e. where the *t*-value on a 2-sided test has *p*< .05. Based on considerations from pilot data, we planned to transform RT data from the RDT task by log-transformation before computing *t*-values.

@@

For the reliability assessment, we checked LIs for normality using a Shapiro-Wilk test. As the LIs were non-normal, we used a series of Spearman’s correlations to examine the correlation of coefficients for session 1 versus session 2. For tasks where the test-retest reliability was above .75 with the existing test length, we reanalysed the data using just the first N trials, reducing N until reliability becomes unsatisfactory (below *r*= .65). This will allow us to identify the most efficient way to administer these tasks in future, while retaining good reliability. However, the full data (including all trials) was used for all subsequent analyses. To address hypothesis 1 and examine the laterality of each task at the population level, we conducted a series of one-sample *t*-tests for each task using session 1 data.
